---
title: "Flight metrics - Black-bellied Sandgrouse (Pterocles orientalis)"
author: "Tiago Crispim-Mendes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}

project_root <- Sys.getenv("FLIGHT_METRICS_PROJECT_ROOT")

if (!nzchar(project_root)) {
  stop(
    "Set the FLIGHT_METRICS_PROJECT_ROOT environment variable ",
    "to the local project directory before rendering this document."
  )
}

if (!dir.exists(project_root)) {
  stop("FLIGHT_METRICS_PROJECT_ROOT does not point to an existing directory.")
}

knitr::opts_knit$set(root.dir = project_root)

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

```

# Load packages

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(lubridate)
library(sf)
library(sp)
library(amt)
library(momentuHMM)
library(lookup)
library(raster)
library(DT)
library(htmltools)
library(future)
library(furrr)
library(purrr)

# Compute great-circle distance between successive locations in kilometres
haversine_distance <- function(lat1, lon1, lat2, lon2) {
  r <- 6371  # Earth's radius in km
  p <- pi / 180
  a <- 0.5 - cos((lat2 - lat1) * p) / 2 +
    cos(lat1 * p) * cos(lat2 * p) *
    (1 - cos((lon2 - lon1) * p)) / 2
  return(2 * r * asin(sqrt(a)))  # result in km
}

```

# Paths

Define input and output directories used throughout the workflow

```{r}

project_data_dir <- "Data"
project_results_dir <- "Results"

species_dir <- file.path("Black_bellied_sandgrouse")
raw_data_dir <- file.path(project_data_dir, species_dir, "Raw")
derived_data_dir <- file.path(project_data_dir, species_dir, "Data_paper")
results_dir <- file.path(project_results_dir, species_dir, "Data_paper")

# Create output directories if they do not exist
dir.create(derived_data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

# Use Pin-tailed sandgrouse boost data for flight-distance calibration
calibration_species_dir <- file.path("Pin_tailed_sandgrouse")
calibration_derived_data_dir <- file.path(project_data_dir, calibration_species_dir, "Data_paper")

```

# Load individual information

Load metadata used to match device identifiers to individuals and retrieve deployment, sex, age, and device information

```{r}

individual_info <- read.csv(
  file.path(
    raw_data_dir,
    "Individuals_data",
    "Info_individuals.csv"
  ),
  sep = ";",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE
)

```

# Data preparation

Prepare per-individual tracking datasets, identify acquisition modes, derive phenological phase and daytime period, extract terrain elevation, and create the 30-min resampled sequences used in later analyses

```{r, warning=FALSE}

# List all raw tracking files
files <- list.files(path = raw_data_dir, pattern = "\\.csv$")

# Define speed thresholds used to identify flight records and valid boost flights
flight_speed_threshold_km_h <- 7
displacement_speed_threshold_km_h <- 3

# Load the UTM grid used to match each location to the appropriate DEM tile
utm_grid <- st_read(
  "MDT/MTN25_ACTUAL_ETRS89_Peninsula_Baleares_Canarias.shp",
  quiet = TRUE
)
utm_grid <- st_transform(utm_grid, CRS("+init=epsg:25830"))

# Define the maximum interval used to group consecutive high-frequency fixes into boost sequences
boost_time_threshold_sec <- 30

# Store one summary row per individual
individual_summary <- NULL

for (i in files) {
  tryCatch({
    
    # Read one raw tracking file and preserve chronological order
    x <- read.csv(file.path(raw_data_dir, i), sep = ",")
    x <- x[rowSums(is.na(x)) != ncol(x), ]
    
    x$UTC_datetime <- as.POSIXct(
      x$Collecting.time,
      format = "%Y-%m-%dT%H:%M:%SZ",
      tz = "GMT"
    )
    
    x <- x %>%
      dplyr::arrange(UTC_datetime)
    
        # Harmonise device identifier for downstream joins and summaries
    x$device_id <- as.character(x$UUID)
    x <- x %>% dplyr::select(-UUID)
    
    # Match each file to a single tracked individual using device_id
    individual <- lookup(
      unique(na.omit(x$device_id)),
      individual_info$IDnum,
      individual_info$ID_new
    )
    x$Individual_ID <- individual
    
    # Shift exact midnight timestamps slightly to avoid ordering problems
    x <- x %>%
      mutate(
        UTC_datetime = if_else(
          format(UTC_datetime, "%H:%M:%S") == "00:00:00",
          UTC_datetime + seconds(1),
          UTC_datetime
        )
      )
    
    # Derive a date field used in filtering and summary outputs
    x$UTC_date <- as.Date(x$UTC_datetime)
    
    # Harmonise altitude fields across files
    if ("Altitude" %in% names(x)) {
      x <- x %>%
        dplyr::rename(Altitude_m = Altitude)
    }
    
    if ("Altitude..Ellipsoid." %in% names(x)) {
      x <- x %>%
        dplyr::rename(Altitude_ellipsoid_m = `Altitude..Ellipsoid.`)
    }
    
    # Harmonise speed and satellite-count field names across files
    if ("Speed" %in% names(x)) {
      x <- x %>%
        dplyr::rename(speed_km_h = Speed)
    }
    
    if ("Satellite.used" %in% names(x)) {
      x <- x %>%
        dplyr::rename(satcount = `Satellite.used`)
    }
    
    # Remove raw fields that are not used in the cleaned workflow
    x <- x %>%
      dplyr::select(-dplyr::any_of(c(
        "Data.source",
        "Data.Source"
      )))
    
    # Exclude the first week after deployment and all records after the last transmission
    deploy <- lookup(individual, individual_info$ID_new, individual_info$capture_date)
    death <- lookup(individual, individual_info$ID_new, individual_info$last_location)
    
    x <- x[x$UTC_date > as.Date(deploy) + 7, ]
    x <- x[x$UTC_date < as.Date(death), ]
    
    # Create a stable row identifier for joins throughout the workflow
    x$IDorder <- seq_len(nrow(x))
    
    # Append device metadata to each record
    x$gps_brand <- lookup(individual, individual_info$ID_new, individual_info$GPS_Brand)
    x$gps_model <- lookup(individual, individual_info$ID_new, individual_info$GPS_Model)
    x$gps_model_full<-lookup(individual,individual_info$ID_new,individual_info$GPS_Model_full)

    
    # Derive coordinates in geographic and projected reference systems
    x_sf <- st_as_sf(x, coords = c("Longitude", "Latitude"), crs = 4326)
    
    x <- x_sf %>%
      mutate(
        X = st_coordinates(.)[, 1],
        Y = st_coordinates(.)[, 2],
        X_25829 = st_coordinates(st_transform(geometry, crs = 25829))[, 1],
        Y_25829 = st_coordinates(st_transform(geometry, crs = 25829))[, 2],
        X_25830 = st_coordinates(st_transform(geometry, crs = 25830))[, 1],
        Y_25830 = st_coordinates(st_transform(geometry, crs = 25830))[, 2]
      ) %>%
      st_drop_geometry()
    
    # Assign each record to the focal phenological phases used in the analyses
    classify_phenology <- function(x) {
      x <- x %>%
        mutate(
          phen_phase = case_when(
            (month(UTC_datetime) == 6 & day(UTC_datetime) >= 1) |
              (month(UTC_datetime) %in% 7:8) |
              (month(UTC_datetime) == 9 & day(UTC_datetime) <= 30) ~ "breeding",
            (month(UTC_datetime) == 11 & day(UTC_datetime) >= 1) |
              (month(UTC_datetime) == 12) |
              (month(UTC_datetime) == 1) |
              (month(UTC_datetime) == 2 & day(UTC_datetime) <= 29) ~ "winter",
            TRUE ~ "other"
          )
        )
      return(x)
    }
    
    x <- classify_phenology(x)
    
    # Calculate sunrise and sunset for each GPS fix
    sun_input <- x %>%
      transmute(
        IDorder,
        date = as.Date(UTC_date),
        lat = Y,
        lon = X
      )
    
    sun_times <- suncalc::getSunlightTimes(
      data = sun_input %>%
        dplyr::select(date, lat, lon),
      keep = c("sunrise", "sunset"),
      tz = "UTC"
    ) %>%
      mutate(
        IDorder = sun_input$IDorder
      ) %>%
      dplyr::select(
        IDorder,
        sunrise,
        sunset
      )
    
    # Classify each GPS fix by daytime period using custom dawn/dusk windows
    x <- x %>%
      left_join(sun_times, by = "IDorder") %>%
      mutate(
        tod = case_when(
          UTC_datetime >= sunrise - hours(1) &
            UTC_datetime <= sunrise + hours(1) ~ "dawn",
          UTC_datetime >= sunset - hours(1) &
            UTC_datetime <= sunset + hours(1) ~ "dusk",
          UTC_datetime > sunrise + hours(1) &
            UTC_datetime < sunset - hours(1) ~ "day",
          TRUE ~ "night"
        ),
        tod = factor(
          tod,
          levels = c("dawn", "day", "dusk", "night")
        )
      ) %>%
      dplyr::select(-sunrise, -sunset)
    
    # Compute elapsed time between consecutive records
    x$lag_original <- c(NA, diff(x$UTC_datetime, units = "secs"))
    
    # Identify high-frequency sequences corresponding to boost acquisitions
    x$burst <- ifelse(x$lag_original < boost_time_threshold_sec, 1, 0)
    
    x$burst2 <- NA
    x$acquisition_mode <- NA
    x <- x[!is.na(x$UTC_datetime), ]
    rownames(x) <- 1:nrow(x)
    
    y <- rle(x$burst == 1)
    y$sumcum <- cumsum(y$lengths)
    
    for (j in 2:length(y$sumcum)) {
      
      if (y$values[j] == "TRUE") {
        
        ya <- y$sumcum[j - 1]
        yb <- y$sumcum[j]
        
        x$burst2[ya:yb] <- y$lengths[j] + 1
        x$acquisition_mode[ya:yb] <- "boost"
        x$acquisition_mode[yb] <- "boost_end"
        
      } else {
        
        if (y$lengths[j] == 1) {
        } else {
          
          ya <- y$sumcum[j - 1] + 1
          yb <- y$sumcum[j] - 1
          
          x$burst2[ya:yb] <- y$lengths[j] - 1
          x$acquisition_mode[ya:yb] <- "simple"
        }
      }
    }
    
    # Reduce acquisition labels to the categories used downstream
    x$acquisition_mode2 <- x$acquisition_mode
    x$acquisition_mode2[x$acquisition_mode == "boost" & x$burst == 0] <- "boost_start"
    
    x$acquisition_mode <- x$acquisition_mode2
    x$acquisition_mode2 <- NULL
    
    # Use speed to separate clearly stationary records from likely flight records
    x <- x %>%
      mutate(
        behavior = case_when(
          speed_km_h == 0 ~ "stationary",
          speed_km_h > flight_speed_threshold_km_h ~ "flight",
          TRUE ~ NA_character_
        )
      )
    
    # Flag low-quality points and implausible displacements before further processing
    x$filter <- ifelse(
      x$X == 0 |
        x$Y == 0 |
        is.na(x$satcount) |
        x$satcount < 4 |
        x$satcount == -99999,
      "REMOVE",
      "OK"
    )
    
    data <- x %>%
      filter(
        acquisition_mode %in% c("simple", "boost_start") &
          filter == "OK"
      ) %>%
      mutate(
        distancefromKM = haversine_distance(lag(Y), lag(X), Y, X)
      )
    
    data$difftimefrom <- c(NA, diff(data$UTC_datetime, units = "secs"))
    data$difftimefromHR <- data$difftimefrom / 3600
    data$KMH <- as.numeric(data$distancefromKM / data$difftimefromHR)
    data$outlier <- ifelse(data$KMH > 150, "yes", "no")
    
    x$outlier <- lookup(x$IDorder, data$IDorder, data$outlier)
    
    x$filter <- ifelse(
      x$outlier == "yes" |
        x$X == 0 |
        x$Y == 0 |
        is.na(x$satcount) |
        x$satcount < 4 |
        x$satcount == -99999,
      "REMOVE",
      "OK"
    )
    
    # Extract ground elevation from the appropriate DEM tile and derive altitude above ground
    altitude_input_data <- x %>%
      dplyr::filter(is.na(filter) | filter == "OK")
    
    altitude_input_data$X_25830_original <- altitude_input_data$X_25830
    altitude_input_data$Y_25830_original <- altitude_input_data$Y_25830
    
    altitude_points_sf <- st_as_sf(
      x = altitude_input_data,
      coords = c("X_25830", "Y_25830"),
      crs = CRS("+init=epsg:25830")
    )
    
    altitude_points_intersection <- st_intersection(utm_grid, altitude_points_sf)
    
    altitude_points_intersection <- altitude_points_intersection[
      , !(names(altitude_points_intersection) %in% c("gid", "huso2", "solape", "edicion"))
    ]
    
    altitude_points_intersection$X_25830 <- altitude_points_intersection$X_25830_original
    altitude_points_intersection$Y_25830 <- altitude_points_intersection$Y_25830_original
    
    # Store terrain extraction results across DEM tiles
    altitude_results_list <- list()
    
    for (j in unique(na.omit(altitude_points_intersection$numero))) {
      tryCatch({
        altitude_points_zone <- altitude_points_intersection[altitude_points_intersection$numero == j, ]
        altitude_points_zone <- data.frame(altitude_points_zone)
        altitude_points_zone <- altitude_points_zone[!is.na(altitude_points_zone$X_25830), ]
        altitude_points_zone <- altitude_points_zone[, !(names(altitude_points_zone) %in% c("geometry"))]
        
        coordinates(altitude_points_zone) <- cbind(
          altitude_points_zone$X_25830,
          altitude_points_zone$Y_25830
        )
        crs(altitude_points_zone) <- CRS("+init=epsg:25830")
        
        mdt_file <- NULL
        
        # Extract terrain elevation using the appropriate UTM zone
        altitude_points_zone$ground_elevation <- ifelse(
          altitude_points_zone$huso == 30,
          {
            mdt_file <- paste0(
              "MDT/MDT02-ETRS89-HU",
              unique(na.omit(altitude_points_zone$huso)),
              "-",
              j,
              "-COB2.tif"
            )
            mdt <- raster(mdt_file)
            crs(mdt) <- CRS("+init=epsg:25830")
            raster::extract(mdt, altitude_points_zone)
          },
          ifelse(
            altitude_points_zone$huso == 29,
            {
              altitude_points_zone <- spTransform(
                altitude_points_zone,
                CRS("+init=epsg:25829")
              )
              
              mdt_file <- paste0(
                "MDT/MDT02-ETRS89-HU",
                unique(na.omit(altitude_points_zone$huso)),
                "-",
                j,
                "-COB2.tif"
              )
              mdt <- raster(mdt_file)
              crs(mdt) <- CRS("+init=epsg:25829")
              raster::extract(mdt, altitude_points_zone)
            },
            NA
          )
        )
        
        altitude_points_zone <- as.data.frame(altitude_points_zone)
        altitude_points_zone <- altitude_points_zone[
          , !(names(altitude_points_zone) %in% c("coords.x1", "coords.x2"))
        ]
        
        altitude_points_zone$altitude_above_ground <-
          altitude_points_zone$Altitude_m - altitude_points_zone$ground_elevation
        
        altitude_results_list[[length(altitude_results_list) + 1]] <- altitude_points_zone
        
      }, error = function(e) {
        cat(
          "ERROR:",
          individual,
          conditionMessage(e),
          if (!is.null(mdt_file)) mdt_file else "",
          "\n"
        )
      })
    }
    
    altitude_results <- dplyr::bind_rows(altitude_results_list) %>%
      dplyr::select(dplyr::any_of(c("IDorder", "huso", "ground_elevation", "altitude_above_ground")))
    
    # Merge terrain-based altitude estimates back into the full dataset
    if (nrow(altitude_results) == 0) {
      x$huso <- NA
      x$ground_elevation <- NA
      x$altitude_above_ground <- NA
    } else {
      x <- x %>%
        left_join(altitude_results, by = "IDorder")
    }
    
    # Assign UTM zone 29 to unmatched records after DEM extraction
    if (any(is.na(x$huso))) {
      x$huso[is.na(x$huso)] <- 29
    }
    
    # Assign a boost identifier to each boost sequence
    x$boostID <- NA
    id <- 1
    for (k in 2:nrow(x)) {
      tryCatch({
        if (!is.na(x$acquisition_mode[k]) && x$acquisition_mode[k] == "boost_start") {
          x2 <- x[k:nrow(x), ]
          x$boostID[k:rownames(x2[x2$acquisition_mode == "boost_end", ])[1]] <- paste("boost_", id, sep = "")
          id <- id + 1
        }
      }, error = function(e) { cat("ERROR :", conditionMessage(e), "\n") })
    }
    
    # Identify boost sequences that likely correspond to flight events
    x$KMH_boost <- NA
    x$boost_flight_ID <- NA
    flight_id <- 1
    
    for (k in 2:nrow(x)) {
      tryCatch({
        if (!is.na(x$acquisition_mode[k]) && x$acquisition_mode[k] == "boost_start") {
          
          x2 <- x[k:nrow(x), ]
          boost_end_idx <- rownames(x2[x2$acquisition_mode == "boost_end", ])[1]
          
          if (!is.na(boost_end_idx)) {
            boost_range <- k:as.integer(boost_end_idx)
            
            speed_condition <- any(x$speed_km_h[boost_range] > flight_speed_threshold_km_h, na.rm = TRUE)
            
            start_point <- x[k, ]
            end_point <- x[as.integer(boost_end_idx), ]
            
            distance_boost <- haversine_distance(start_point$Y, start_point$X, end_point$Y, end_point$X)
            
            time_boost_sec <- as.numeric(difftime(end_point$UTC_datetime, start_point$UTC_datetime, units = "secs"))
            
            KMH_boost <- ifelse(time_boost_sec > 0, (distance_boost / time_boost_sec) * 3600, 0)
            
            x$KMH_boost[boost_range] <- KMH_boost
            
            if (speed_condition && KMH_boost > displacement_speed_threshold_km_h) {
              x$boost_flight_ID[boost_range] <- paste("boostflight_", flight_id, sep = "")
              flight_id <- flight_id + 1
            }
          }
        }
      }, error = function(e) { cat("ERROR :", conditionMessage(e), "\n") })
    }
    
    # Resample valid records to a standard 30-min interval within each phenological phase
    x_split <- x %>%
      filter(!is.na(filter) & filter == "OK") %>%
      group_split(phen_phase)
    
    x_processed <- lapply(x_split, function(df) {
      
      x_track_data <- make_track(
        df, .x = X, .y = Y, .t = UTC_datetime,
        crs = 4326, IDorder = IDorder
      )
      
      x_resampled_track <- x_track_data |>
        track_resample(rate = minutes(30), tolerance = minutes(2)) |>
        filter_min_n_burst(min_n = 3)
      
      df <- df %>%
        left_join(
          dplyr::select(x_resampled_track, IDorder, ID_keep = burst_),
          by = "IDorder"
        )
      
      # Store each retained 30-min sequence as an individual- and phase-specific identifier
      df$ID_keep <- ifelse(
        !is.na(df$ID_keep),
        paste0(df$Individual_ID, "_", df$phen_phase, "_", df$ID_keep),
        df$ID_keep
      )
      df$ID_keep <- as.character(df$ID_keep)
      
      return(df)
    })
    
    all_processed <- bind_rows(x_processed)
    
    x <- x %>%
      left_join(
        dplyr::select(all_processed, IDorder, ID_keep),
        by = "IDorder"
      )
    
    # Summarise tracking coverage and the amount of data retained at each step
    individual_summary_temp <- tibble(
      individual_id = unique(x$Individual_ID),
      sex = if ("Sex" %in% names(individual_info)) {
        lookup(individual, individual_info$ID_new, individual_info$Sex)
      } else {
        NA_character_
      },
      age = if ("Age" %in% names(individual_info)) {
        lookup(individual, individual_info$ID_new, individual_info$Age)
      } else {
        NA_character_
      },
      gps_brand = unique(x$gps_brand),
      gps_model = unique(x$gps_model),
      gps_model_full = unique(x$gps_model_full),
      tracking_start = min(x$UTC_date, na.rm = TRUE),
      tracking_end = max(x$UTC_date, na.rm = TRUE),
      tracking_duration_days = round(as.numeric(
        difftime(
          max(x$UTC_datetime, na.rm = TRUE),
          min(x$UTC_datetime, na.rm = TRUE),
          units = "days"
        )
      )),
      n_fixes_raw = nrow(x),
      n_fixes_30min = sum(!is.na(x$ID_keep)),
      n_fixes_stationary = sum(x$behavior == "stationary", na.rm = TRUE),
      n_fixes_flight = sum(x$behavior == "flight", na.rm = TRUE)
    )
    
    phases <- c("winter", "breeding", "other")
    
    # Add phase-specific counts for the main acquisition and behavioural categories
    for (phase in phases) {
      x_phase <- x[x$phen_phase == phase & !is.na(x$phen_phase), ]
      individual_summary_temp[[paste0("n_fixes_simple_", phase)]] <- sum(x_phase$acquisition_mode == "simple", na.rm = TRUE)
      individual_summary_temp[[paste0("n_burst_events_", phase)]] <- sum(x_phase$acquisition_mode == "burst_end", na.rm = TRUE)
      individual_summary_temp[[paste0("n_boost_events_", phase)]] <- sum(x_phase$acquisition_mode == "boost_end", na.rm = TRUE)
      individual_summary_temp[[paste0("n_boost_flight_events_", phase)]] <- length(unique(na.omit(x_phase$boost_flight_ID)))
      individual_summary_temp[[paste0("n_fixes_30min_", phase)]] <- sum(!is.na(x_phase$ID_keep))
      individual_summary_temp[[paste0("n_sequences_30min_", phase)]] <- length(unique(na.omit(x_phase$ID_keep)))
      individual_summary_temp[[paste0("n_fixes_stationary_", phase)]] <- sum(x_phase$behavior == "stationary" & !is.na(x_phase$altitude_above_ground), na.rm = TRUE)
      individual_summary_temp[[paste0("n_fixes_flight_", phase)]] <- sum(x_phase$behavior == "flight" & !is.na(x_phase$altitude_above_ground), na.rm = TRUE)
    }
    
    individual_summary <- rbind(individual_summary, individual_summary_temp)
    
    # Save the prepared per-individual dataset used in downstream analyses
    write.csv(x, file.path(derived_data_dir, paste0(individual, "_Data_Prep.csv")), row.names = FALSE)
    
  }, error = function(e) { cat("ERROR:", individual, conditionMessage(e), "\n") })
}

# Save the individual-level summary table
individual_summary <- individual_summary %>% arrange(individual_id)
write.csv(individual_summary, file.path(results_dir, "Individuals_data_summary.csv"), row.names = FALSE)

tagList(
  h3("Individual summary"),
  datatable(individual_summary, options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 10))
)

```

# Boosts

## Correction factor for flight distances

Summarise high-frequency boost flights to estimate the ratio between cumulative path length and straight-line displacement

```{r}

# List prepared per-individual datasets
files <- list.files(path = calibration_derived_data_dir, pattern = "Data_Prep\\.csv$")

# Store boost records and per-flight summaries across all individuals
boost_flight_data_list <- list()
boost_distance_summary_list <- list()

for (i in files) {
  tryCatch({
    # Read one prepared dataset and preserve chronological order
    df_complete <- read.csv(file.path(calibration_derived_data_dir, i), sep = ",") %>%
      dplyr::mutate(
        Individual_ID = as.character(Individual_ID),
        device_id = as.character(device_id),
        boost_flight_ID = as.character(boost_flight_ID),
        boostID = as.character(boostID),
        phen_phase = as.character(phen_phase),
        acquisition_mode = as.character(acquisition_mode),
        behavior = as.character(behavior),
        filter = as.character(filter)
      )
    
    df_complete$UTC_datetime <- as.POSIXct(
      df_complete$UTC_datetime,
      format = "%Y-%m-%d %H:%M:%OS",
      tz = "GMT"
    )
    
    df_complete <- df_complete[order(df_complete$UTC_datetime), ]
    
    # Keep only boost sequences previously classified as flight
    df_filtered <- df_complete[!is.na(df_complete$boost_flight_ID), ]
    
    if (nrow(df_filtered) > 0) {
      boost_flight_data_list[[length(boost_flight_data_list) + 1]] <- df_filtered
    }
    
    # Summarise each boost flight independently
    for (boost_id in unique(df_filtered$boost_flight_ID)) {
      
      df_boost <- df_filtered[df_filtered$boost_flight_ID == boost_id, ]
      df_boost <- df_boost[order(df_boost$UTC_datetime), ]
      
      # Skip incomplete boost flights with fewer than two records
      if (nrow(df_boost) < 2) next
      
      phen_phase <- unique(df_boost$phen_phase)[1]
      
      # Sum successive step lengths within the boost sequence
      df_boost$distancefromKM <- haversine_distance(
        lag(df_boost$Y),
        lag(df_boost$X),
        df_boost$Y,
        df_boost$X
      )
      
      total_distance <- sum(df_boost$distancefromKM, na.rm = TRUE)
      
      # Compare cumulative distance with straight-line displacement
      first_point <- df_boost[1, ]
      last_point <- df_boost[nrow(df_boost), ]
      
      displacement <- haversine_distance(
        first_point$Y,
        first_point$X,
        last_point$Y,
        last_point$X
      )
      
      correction_factor <- ifelse(displacement > 0, total_distance / displacement, NA_real_)
      
      result <- data.frame(
        Individual_ID = unique(df_boost$Individual_ID)[1],
        boost_flight_ID = boost_id,
        first_datetime = first_point$UTC_datetime,
        last_datetime = last_point$UTC_datetime,
        phen_phase = phen_phase,
        total_distance_Km = total_distance,
        displacement_Km = displacement,
        correction_factor = correction_factor
      )
      
      boost_distance_summary_list[[length(boost_distance_summary_list) + 1]] <- result
    }
    
  }, error = function(e) {
    cat("ERROR in file:", i, "-", conditionMessage(e), "\n")
  })
}

# Combine all boost-level outputs
boost_flight_data <- dplyr::bind_rows(boost_flight_data_list)
boost_distance_summary <- dplyr::bind_rows(boost_distance_summary_list)

# Summarise correction factors by phenological phase
summary_table_correction_factor <- boost_distance_summary %>%
  group_by(phen_phase) %>%
  summarise(
    mean_correction_factor = mean(correction_factor, na.rm = TRUE),
    sd_correction_factor = sd(correction_factor, na.rm = TRUE),
    median_correction_factor = median(correction_factor, na.rm = TRUE)
  )

# Add a global summary across all phases
total_summary_correction_factor <- boost_distance_summary %>%
  summarise(
    phen_phase = "Global",
    mean_correction_factor = mean(correction_factor, na.rm = TRUE),
    sd_correction_factor = sd(correction_factor, na.rm = TRUE),
    median_correction_factor = median(correction_factor, na.rm = TRUE)
  )

summary_correction_factor <- bind_rows(summary_table_correction_factor, total_summary_correction_factor)

tagList(
  h3("Correction factor"),
  datatable(summary_correction_factor, options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 5))
)

```

## Boost displacement 30 min

Match each boost start to the nearest valid location approximately 30 min later to derive the flight displacement used to inform HMM initial values

```{r, warning=FALSE}

# List prepared per-individual datasets
files <- list.files(path = calibration_derived_data_dir, pattern = "Data_Prep\\.csv$")

# Store one summary row per matched boost pair
boost_displacement_summary <- NULL

for (file_name in files) {
  tryCatch({
    # Read one prepared dataset and preserve chronological order
    df_complete <- read.csv(
      file.path(calibration_derived_data_dir, file_name),
      sep = ","
    )
    
    df_complete$UTC_datetime <- as.POSIXct(
      df_complete$UTC_datetime,
      format = "%Y-%m-%d %H:%M:%OS",
      tz = "GMT"
    )
    
    df_complete <- df_complete[order(df_complete$UTC_datetime), ]
    
    # Use only valid records when matching the 30-min endpoint
    df_filtered <- df_complete[!is.na(df_complete$filter) & df_complete$filter == "OK", ]
    
    # Keep boost starts associated with boost flights
    df_boost <- df_filtered %>%
      dplyr::filter(acquisition_mode == "boost_start", !is.na(boost_flight_ID))
    
    if (nrow(df_boost) > 0) {
      
      # Store matched start-end pairs
      df_boost_pairs <- data.frame()
      counter <- 1
      
      for (j in 1:nrow(df_boost)) {
        start_point <- df_boost[j, ]
        target_time <- start_point$UTC_datetime + 1800
        
        # Search for the nearest valid point after the boost start
        candidates <- df_filtered %>%
          dplyr::filter(UTC_datetime > start_point$UTC_datetime)
        
        if (nrow(candidates) > 0) {
          candidates$delta_abs <- abs(as.numeric(
            difftime(candidates$UTC_datetime, target_time, units = "secs")
          ))
          
          closest_candidate <- candidates[which.min(candidates$delta_abs), ]
          
          delta_real <- as.numeric(
            difftime(closest_candidate$UTC_datetime, start_point$UTC_datetime, units = "secs")
          )
          
          # Retain matches within a ±3 min tolerance around 30 min
          if (delta_real >= 1620 && delta_real <= 1980) {
            temp_rows <- dplyr::bind_rows(start_point, closest_candidate)
            temp_rows$ID_boost_to <- paste0(start_point$device_id, "_", counter)
            df_boost_pairs <- dplyr::bind_rows(df_boost_pairs, temp_rows)
            counter <- counter + 1
          }
        }
      }
      
      # Summarise each matched pair as a single 30-min displacement value
      if (nrow(df_boost_pairs) > 0) {
        
        df_boost_pairs <- df_boost_pairs %>%
          dplyr::arrange(ID_boost_to, UTC_datetime) %>%
          dplyr::group_by(ID_boost_to) %>%
          dplyr::mutate(
            displacement_30min = haversine_distance(dplyr::lag(Y), dplyr::lag(X), Y, X)
          ) %>%
          dplyr::summarise(
            Individual_ID = dplyr::first(Individual_ID),
            boost_flight_ID = dplyr::first(stats::na.omit(boost_flight_ID)),
            phen_phase = dplyr::first(phen_phase),
            displacement_30min = dplyr::first(stats::na.omit(displacement_30min)),
            .groups = "drop"
          )
        
        boost_displacement_summary <- dplyr::bind_rows(
          boost_displacement_summary,
          df_boost_pairs
        )
      }
    }
  }, error = function(e) { cat("ERROR:", file_name, "-", conditionMessage(e), "\n") })
}

# Summarise 30-min displacement globally and by phenological phase
flight_displacement_calibration_global <- boost_displacement_summary %>%
  summarise(
    phen_phase = "Global",
    mean_displacement = mean(displacement_30min, na.rm = TRUE),
    sd_displacement = sd(displacement_30min, na.rm = TRUE)
  )

flight_displacement_calibration_by_phase <- boost_displacement_summary %>%
  group_by(phen_phase) %>%
  summarise(
    mean_displacement = mean(displacement_30min, na.rm = TRUE),
    sd_displacement = sd(displacement_30min, na.rm = TRUE)
  )

flight_displacement_summary <- bind_rows(
  flight_displacement_calibration_global,
  flight_displacement_calibration_by_phase
)

tagList(
  h3("Flight displacement summary"),
  datatable(flight_displacement_summary, options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 5))
)

```

## Boost flight data

Derive step lengths and turning angles for boost flights and export the boost-level movement dataset used for calibration

```{r, warning=FALSE}

# boost_flight_data <- boost_flight_data %>%
#   mutate(
#     ID = paste0(Individual_ID, "_", boost_flight_ID),
#     x = as.numeric(X),
#     y = as.numeric(Y),
#     time = as.POSIXct(UTC_datetime, tz = "UTC")
#   ) %>%
#   filter(!is.na(ID), !is.na(x), !is.na(y), !is.na(time)) %>%
#   arrange(ID, time)
# 
# boost_flight_data_step_angle <- prepData(boost_flight_data, type = "LL")
# 
# # Export the boost-level dataset with step lengths and turning angles
# boost_flight_data_filtered <- boost_flight_data_step_angle %>%
#   dplyr::select(
#     Individual_ID,
#     boost_flight_ID,
#     step,
#     angle,
#     UTC_datetime,
#     phen_phase,
#     tod,
#     acquisition_mode,
#     satcount,
#     U_bat_mV,
#     bat_soc_pct,
#     solar_I_mA,
#     hdop,
#     Altitude_m,
#     speed_km_h,
#     direction_deg,
#     temperature_C,
#     mag_x, mag_y, mag_z,
#     acc_x, acc_y, acc_z,
#     behavior,
#     ground_elevation,
#     altitude_above_ground
#   )
# 
# write.csv(boost_flight_data_filtered,file.path(results_dir,"Boost_flight_data.csv"),row.names = FALSE)

```

## Boost summary

Combine within-boost path metrics with the matched 30-min displacement summary

```{r, warning=FALSE}

# boost_flight_summary <- boost_distance_summary %>%
#   left_join(
#     boost_displacement_summary,
#     by = c("Individual_ID", "boost_flight_ID", "phen_phase")
#   )%>%
#   dplyr::select(-ID_boost_to)
# 
# 
# write.csv(boost_flight_summary,file.path(results_dir,"Boost_flight_summary.csv"),row.names = FALSE)
# 
# tagList(
#   h3("Boost flight summary"),
#   datatable(boost_flight_summary, options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 5))
# )

```

# Flight Distance

## HMM

### Data preparation

Build the HMM input datasets from 30-min sequences and derive step lengths and turning angles separately for breeding and winter

```{r, warning=FALSE}

# List prepared per-individual datasets
files <- list.files(path = derived_data_dir,pattern = "Data_Prep\\.csv$")

# Store all retained 30-min records across individuals
data_list <- list()

for (i in files) {
  df_complete <- read.csv(
    file.path(derived_data_dir, i),
    sep = ","
  )
  
  # Retain only records belonging to 30-min sequences
  df_filtered <- df_complete %>%
    dplyr::filter(!is.na(ID_keep)) %>%
    dplyr::mutate(
      device_id = as.character(device_id),
      ID_keep = as.character(ID_keep),
      boost_flight_ID = as.character(boost_flight_ID),
      boostID = as.character(boostID))
  
  data_list[[length(data_list) + 1]] <- df_filtered
}

hmm_filtered_data <- dplyr::bind_rows(data_list)

hmm_filtered_data$UTC_datetime <- as.POSIXct(
  hmm_filtered_data$UTC_datetime,
  format = "%Y-%m-%d %H:%M:%OS",
  tz = "GMT"
)

# Prepare the coordinate and time fields required by momentuHMM
hmm_filtered_data <- hmm_filtered_data %>%
  dplyr::mutate(
    ID = ID_keep,
    x = as.numeric(X),
    y = as.numeric(Y),
    time = UTC_datetime
  ) %>%
  dplyr::filter(!is.na(ID), !is.na(x), !is.na(y), !is.na(time))

# Fit separate HMM inputs for the two focal phenological phases
hmm_filtered_data_breeding <- hmm_filtered_data %>%
  dplyr::filter(phen_phase == "breeding") %>%
  dplyr::arrange(ID, time)

hmm_filtered_data_winter <- hmm_filtered_data %>%
  dplyr::filter(phen_phase == "winter") %>%
  dplyr::arrange(ID, time)

hmm_filtered_data_other <- hmm_filtered_data %>%
  dplyr::filter(phen_phase == "other") %>%
  dplyr::arrange(ID, time)

# Derive step length and turning angle for each sequence
hmm_data_breeding <- prepData(hmm_filtered_data_breeding, type = "LL")
hmm_data_breeding$angle[hmm_data_breeding$step == 0] <- 0
hmm_data_breeding$step[hmm_data_breeding$step == 0] <- 0.0001

hmm_data_winter <- prepData(hmm_filtered_data_winter, type = "LL")
hmm_data_winter$angle[hmm_data_winter$step == 0] <- 0
hmm_data_winter$step[hmm_data_winter$step == 0] <- 0.0001

hmm_data_other <- prepData(hmm_filtered_data_other, type = "LL")
hmm_data_other$angle[hmm_data_other$step == 0] <- 0
hmm_data_other$step[hmm_data_other$step == 0] <- 0.0001
write.csv(hmm_data_other, file.path(derived_data_dir, "hmm_data_other.csv"), row.names = FALSE)

```

### breeding

#### Model fitting

Use boost-derived displacement to define biologically plausible starting values for the 3-state HMM fitted to breeding data

```{r fit-hmm-breeding}

# Specify the state-dependent distributions for step length and turning angle
dist <- list(step = "gamma", angle = "vm")

# Use phase-specific displacement when available, otherwise fall back to the global value
mean_displacement_breeding <- ifelse("breeding" %in% flight_displacement_summary$phen_phase,
                                      flight_displacement_summary$mean_displacement[flight_displacement_summary$phen_phase == "breeding"],
                                      flight_displacement_summary$mean_displacement[flight_displacement_summary$phen_phase == "Global"])

sd_displacement_breeding <- ifelse("breeding" %in% flight_displacement_summary$phen_phase,
                                      flight_displacement_summary$sd_displacement[flight_displacement_summary$phen_phase == "breeding"],
                                      flight_displacement_summary$sd_displacement[flight_displacement_summary$phen_phase == "Global"])


# Provide initial values for the 3-state HMM
par0_3_states <- list(step = c(0.01, 0.02, mean_displacement_breeding, 0.01, 0.010 , sd_displacement_breeding), angle = c(0.01, 0.01, 0.01))

hmm_breeding <- fitHMM(hmm_data_breeding, 
               nbStates = 3, 
               dist = dist, 
               retryFits = 0, # 30
               Par0 = par0_3_states)

print(hmm_breeding)

```

#### Model outputs

Decode the most likely state sequence and export state probabilities to support the interpretation of the fitted breeding HMM

```{r}

hmm_data_breeding$viterbi_state <- factor(viterbi(hmm_breeding))
  
sp_breeding <- stateProbs(hmm_breeding)
colnames(sp_breeding) <- paste0("state_prob_", seq_len(ncol(sp_breeding)))

hmm_data_breeding <- cbind(hmm_data_breeding, sp_breeding)

write.csv(hmm_data_breeding, file.path(derived_data_dir, "hmm_data_breeding.csv"), row.names = FALSE)

# Inspect step-length distributions across the inferred states
ggplot(hmm_data_breeding, aes(x = step, color = viterbi_state, fill = viterbi_state)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Step Length Density by State - breeding",
    x = "Step length (km)",
    y = "Density"
  ) +
  theme_classic() +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)
  )

ggplot(hmm_data_breeding, aes(x = viterbi_state, y = step, fill = viterbi_state)) +
  geom_boxplot() +
   labs(
    title = "Step Length Distribution by State - breeding",
    x = "Movement state",
    y = "Step length (km)"
  ) +
  theme_classic() +
  scale_y_continuous(breaks = seq(0, 3, by = 0.5))+
  coord_cartesian(ylim = c(0, 3))

```

#### Plot estimated state-dependent distributions

Inspect the fitted state-dependent distributions and the proportion of time in each state

```{r, fig.keep = 1:10}

plot(hmm_breeding, breaks = 25, ask = FALSE, plotTracks = FALSE)

timeInStates(hmm_breeding)

```

### winter

#### Model fitting

Use boost-derived displacement to define biologically plausible starting values for the 3-state HMM fitted to winter data

```{r fit-hmm-winter}

dist <- list(step = "gamma", angle = "vm")

mean_displacement_winter <- ifelse("winter" %in% flight_displacement_summary$phen_phase,
                                      flight_displacement_summary$mean_displacement[flight_displacement_summary$phen_phase == "winter"],
                                      flight_displacement_summary$mean_displacement[flight_displacement_summary$phen_phase == "Global"])

sd_displacement_winter <- ifelse("winter" %in% flight_displacement_summary$phen_phase,
                                      flight_displacement_summary$sd_displacement[flight_displacement_summary$phen_phase == "winter"],
                                      flight_displacement_summary$sd_displacement[flight_displacement_summary$phen_phase == "Global"])


# Provide initial values for the 3-state winter HMM
par0_3_states <- list(step = c(0.01, 0.030, mean_displacement_winter, 0.01, 0.010 , sd_displacement_winter), angle = c(0.01, 0.01, 0.01))

hmm_winter <- fitHMM(hmm_data_winter, 
               nbStates = 3, 
               dist = dist, 
               retryFits = 0, # 30
               Par0 = par0_3_states)

print(hmm_winter)

```

#### Model outputs

Decode winter states and export the corresponding state probabilities

```{r}

hmm_data_winter$viterbi_state <- factor(viterbi(hmm_winter))

sp_winter <- stateProbs(hmm_winter)
colnames(sp_winter) <- paste0("state_prob_", seq_len(ncol(sp_winter)))

hmm_data_winter <- cbind(hmm_data_winter, sp_winter)

write.csv(hmm_data_winter, file.path(derived_data_dir, "hmm_data_winter.csv"), row.names = FALSE)

# Inspect step-length distributions across the inferred winter states
ggplot(hmm_data_winter, aes(x = step, color = viterbi_state, fill = viterbi_state)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Step Length Density by State - winter",
    x = "Step length (km)",
    y = "Density"
  ) +
  theme_classic() +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)
  )

ggplot(hmm_data_winter, aes(x = viterbi_state, y = step, fill = viterbi_state)) +
  geom_boxplot() +
   labs(
    title = "Step Length Distribution by State - winter",
    x = "Movement state",
    y = "Step length (km)"
  ) +
  theme_classic() +
  scale_y_continuous(breaks = seq(0, 3, by = 0.5))+
  coord_cartesian(ylim = c(0, 3))

```

#### Plot estimated state-dependent distributions

Inspect the fitted state-dependent distributions and the proportion of time in each state

```{r, fig.keep = 1:10}

plot(hmm_winter, breaks = 25, ask = FALSE, plotTracks = FALSE)

timeInStates(hmm_winter)

```

### Merge phases

Combine the breeding and winter HMM outputs and harmonise the state labels used in the downstream movement summaries

```{r}

hmm_data_breeding<- read.csv(file.path(derived_data_dir, "hmm_data_breeding.csv"), sep = ",")
hmm_data_winter<- read.csv(file.path(derived_data_dir, "hmm_data_winter.csv"), sep = ",")
hmm_data_other <- read.csv(file.path(derived_data_dir, "hmm_data_other.csv"), sep = ",")

data_hmm_all <- bind_rows(hmm_data_breeding, hmm_data_winter, hmm_data_other)

data_hmm_all <- data_hmm_all %>%
  mutate(
    state_label = factor(
      viterbi_state,
      levels = c("1", "2", "3"),
      labels = c("rest", "walk", "fly")
    ),
    phen_phase = factor(phen_phase, levels = c("breeding", "winter", "other"))
  )

# Compare state-dependent step lengths between phases
plot_step_length_distributions <- data_hmm_all %>%
  filter(phen_phase != "other") %>%
  ggplot(aes(x = state_label, y = step, fill = state_label)) +
  geom_boxplot() +
  facet_wrap(~ phen_phase) +
  labs(
    x = "Movement state",
    y = "Step length (km)"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_y_continuous(breaks = seq(0, 3, by = 0.5)) +
  coord_cartesian(ylim = c(0, 3))

print(plot_step_length_distributions)

```

## Calculate daily flight distance  

### Add correction factor to distances

Apply the boost-derived correction factor to HMM step lengths to approximate path length more accurately at the 30-min scale

```{r}

data_hmm_all_corrected <- data_hmm_all %>%
  mutate(
    correction_factor = ifelse(
      viterbi_state == 3,
      ifelse(
        phen_phase %in% summary_correction_factor$phen_phase,
        summary_correction_factor$median_correction_factor[
          match(phen_phase, summary_correction_factor$phen_phase)
        ],
        summary_correction_factor$median_correction_factor[
          summary_correction_factor$phen_phase == "Global"
        ]
      ),
      NA_real_
    ),
    corrected_distance = ifelse(
      viterbi_state == 3,
      step * correction_factor,
      NA_real_
    )
  )

write.csv(data_hmm_all_corrected,file.path(derived_data_dir, "data_hmm_all_corrected.csv"), row.names = FALSE)

# Export the movement dataset used in distance and height summaries
movement_data <- data_hmm_all_corrected %>%
  dplyr::select(
    Individual_ID,
    sequence_id = ID,
    step,
    angle,
    UTC_datetime,
    phen_phase,
    tod,
    acquisition_mode,
    
    ground_elevation,
    altitude_above_ground,
    viterbi_state,	
    state_label,
    state_prob_1,	
    state_prob_2,	
    state_prob_3,
    correction_factor,
    corrected_distance,

    satcount,
    Altitude_m,
    Altitude_ellipsoid_m,
    speed_km_h,
    Course,
    Positioning.mode,
    HDOP,
    VDOP
  )%>%
  mutate(
    acquisition_mode = dplyr::recode(
      acquisition_mode,
      "boost_start" = "boost",
      "boost_end"   = "boost",
      "burst_end"   = "burst"
    )
  )%>%
  arrange(Individual_ID, UTC_datetime)
 
write.csv(movement_data,file.path(results_dir, "Movement_data.csv"), row.names = FALSE)

```

### Descriptive summaries of corrected flight distance by phenological phase and daytime period

```{r}

# Load corrected movement data
data_flight_distance_all <- read.csv(
  file.path(results_dir, "Movement_data.csv"),
  sep = ","
)

# Build continuous 30-min sequences within each original sequence and time-of-day class
sequence_distance_summary <- data_flight_distance_all %>%
  filter(phen_phase != "other") %>%
  mutate(
    UTC_datetime = as.POSIXct(UTC_datetime, tz = "UTC")
  ) %>%
  group_by(sequence_id, tod) %>%
  mutate(
    time_diff = as.numeric(difftime(UTC_datetime, lag(UTC_datetime), units = "mins")),
    new_sequence = ifelse(is.na(time_diff) | time_diff > 35, 1, 0),
    new_sequence_id = cumsum(new_sequence),
    global_sequence_id = paste(sequence_id, tod, new_sequence_id, sep = "_")
  ) %>%
  ungroup() %>%
  group_by(global_sequence_id, phen_phase, tod, Individual_ID) %>%
  summarise(
    hourly_distance_km = sum(corrected_distance[viterbi_state == 3], na.rm = TRUE) / (n() * 0.5),
    n_fixes = dplyr::n(),
    .groups = "drop"
  )

# Descriptive hourly summaries by phenological phase and daytime period
flight_distance_hourly_summary <- sequence_distance_summary %>%
  group_by(phen_phase, tod) %>%
  summarise(
    n_individuals = n_distinct(Individual_ID),
    n_sequences = dplyr::n(),
    mean_hourly_distance_km = mean(hourly_distance_km, na.rm = TRUE),
    sd_hourly_distance_km = sd(hourly_distance_km, na.rm = TRUE),
    median_hourly_distance_km = median(hourly_distance_km, na.rm = TRUE),
    q25_hourly_distance_km = quantile(hourly_distance_km, 0.25, na.rm = TRUE),
    q75_hourly_distance_km = quantile(hourly_distance_km, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    species = "Pterocles orientalis",
    phen_phase = factor(phen_phase, levels = c("breeding", "winter")),
    tod = factor(tod, levels = c("dawn", "day", "dusk", "night"))
  ) %>%
  arrange(phen_phase, tod) %>%
  relocate(species, .before = 1)

tagList(
  h3("Descriptive hourly flight distance by phenological phase and daytime period"),
  datatable(
    flight_distance_hourly_summary,
    options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 8)
  )
)

# Define representative calendar windows for the focal phenological phases
manual_dates <- tibble::tibble(
  phen_phase = c("breeding", "winter"),
  start_date = as.Date(c("2024-06-01", "2024-11-01")),
  end_date   = as.Date(c("2024-09-30", "2025-02-28"))
)

# Load a dataset that still contains geographic coordinates
movement_data_with_coords <- read.csv(
  file.path(derived_data_dir, "data_hmm_all_corrected.csv"),
  sep = ","
)

# Prepare one representative location and three representative dates per phase
phases_days <- manual_dates %>%
  left_join(
    movement_data_with_coords %>%
      filter(phen_phase != "other") %>%
      group_by(phen_phase) %>%
      summarise(
        lon = mean(X, na.rm = TRUE),
        lat = mean(Y, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "phen_phase"
  ) %>%
  mutate(
    mid_date = start_date + floor((end_date - start_date) / 2),
    sim_date = pmap(
      list(start_date, mid_date, end_date),
      ~ c(..1, ..2, ..3)
    )
  ) %>%
  dplyr::select(phen_phase, sim_date, lon, lat) %>%
  tidyr::unnest(sim_date) %>%
  mutate(day_id = row_number())

# Calculate sunrise and sunset for each simulated day
sun_times <- suncalc::getSunlightTimes(
  data = phases_days %>%
    transmute(
      date = sim_date,
      lat = lat,
      lon = lon
    ),
  keep = c("sunrise", "sunset"),
  tz = "UTC"
) %>%
  mutate(day_id = phases_days$day_id) %>%
  dplyr::select(day_id, sunrise, sunset)

# Simulate 30-min points and classify daytime period directly
tod_duration_summary <- phases_days %>%
  left_join(sun_times, by = "day_id") %>%
  tidyr::expand_grid(step = 0:47) %>%
  mutate(
    UTC_datetime = as.POSIXct(sim_date, tz = "UTC") + minutes(30 * step),
    tod = case_when(
      UTC_datetime >= sunrise - hours(1) &
        UTC_datetime <= sunrise + hours(1) ~ "dawn",
      UTC_datetime >= sunset - hours(1) &
        UTC_datetime <= sunset + hours(1) ~ "dusk",
      UTC_datetime > sunrise + hours(1) &
        UTC_datetime < sunset - hours(1) ~ "day",
      TRUE ~ "night"
    ),
    tod = factor(tod, levels = c("dawn", "day", "dusk", "night"))
  ) %>%
  group_by(phen_phase, tod) %>%
  summarise(
    duration_hours = n() * 0.5 / 3,
    .groups = "drop"
  ) %>%
  mutate(
    phen_phase = factor(phen_phase, levels = c("breeding", "winter")),
    tod = factor(tod, levels = c("dawn", "day", "dusk", "night"))
  ) %>%
  arrange(phen_phase, tod)

tagList(
  h3("Average TOD duration by phenological phase"),
  datatable(
    tod_duration_summary,
    options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 8)
  )
)

# Convert hourly descriptive summaries into descriptive daily contributions
flight_distance_daily_tod_summary <- flight_distance_hourly_summary %>%
  left_join(tod_duration_summary, by = c("phen_phase", "tod")) %>%
  mutate(
    mean_daily_distance_km = mean_hourly_distance_km * duration_hours,
    sd_daily_distance_km = sd_hourly_distance_km * duration_hours,
    median_daily_distance_km = median_hourly_distance_km * duration_hours,
    q25_daily_distance_km = q25_hourly_distance_km * duration_hours,
    q75_daily_distance_km = q75_hourly_distance_km * duration_hours
  ) %>%
  arrange(phen_phase, tod)

# Compute one descriptive total daily distance per phenological phase
flight_distance_daily_total <- flight_distance_daily_tod_summary %>%
  group_by(species, phen_phase) %>%
  summarise(
    tod = "total_24h",
    n_individuals = max(n_individuals, na.rm = TRUE),
    n_sequences = sum(n_sequences, na.rm = TRUE),
    mean_hourly_distance_km = NA_real_,
    sd_hourly_distance_km = NA_real_,
    median_hourly_distance_km = NA_real_,
    q25_hourly_distance_km = NA_real_,
    q75_hourly_distance_km = NA_real_,
    duration_hours = 24,
    mean_daily_distance_km = sum(mean_daily_distance_km, na.rm = TRUE),
    sd_daily_distance_km = sqrt(sum(sd_daily_distance_km^2, na.rm = TRUE)),
    median_daily_distance_km = sum(median_daily_distance_km, na.rm = TRUE),
    q25_daily_distance_km = sum(q25_daily_distance_km, na.rm = TRUE),
    q75_daily_distance_km = sum(q75_daily_distance_km, na.rm = TRUE),
    .groups = "drop"
  )

# Combine TOD-specific contributions with the phase-level total
flight_distance_daily_summary <- bind_rows(
  flight_distance_daily_tod_summary,
  flight_distance_daily_total
) %>%
  mutate(
    phen_phase = factor(phen_phase, levels = c("breeding", "winter")),
    tod = factor(tod, levels = c("dawn", "day", "dusk", "night", "total_24h"))
  ) %>%
  arrange(phen_phase, tod)

tagList(
  h3("Descriptive daily flight distance by phenological phase and daytime period"),
  datatable(
    flight_distance_daily_summary,
    options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 10)
  )
)

write.csv(
  flight_distance_daily_summary,
  file.path(results_dir, "Flight_distance_summary.csv"),
  row.names = FALSE
)


plot_flight_distance_descriptive_hourly <- flight_distance_hourly_summary %>%
  ggplot(aes(x = tod, y = mean_hourly_distance_km, fill = phen_phase)) +
  geom_col(position = "dodge") +
  facet_wrap(~ phen_phase) +
  labs(
    x = "Daytime period",
    y = "Mean hourly flight distance (km)"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_fill_manual(
    values = c(
      "breeding" = "#F8766D",
      "winter" = "#619CFF"
    )
  )

plot_flight_distance_descriptive_hourly

ggsave(
  file.path(results_dir, "plot_flight_distance_descriptive_hourly.png"),
  plot_flight_distance_descriptive_hourly,
  width = 15,
  height = 10,
  units = "cm",
  dpi = 600
)

plot_flight_distance_descriptive_daily <- flight_distance_daily_summary %>%
  ggplot(aes(x = tod, y = mean_daily_distance_km, fill = phen_phase)) +
  geom_col(position = "dodge") +
  facet_wrap(~ phen_phase) +
  labs(
    x = "Daytime period",
    y = "Mean daily flight distance (km)"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_x_discrete(
    labels = c(
      dawn = "Dawn",
      day = "Day",
      dusk = "Dusk",
      night = "Night",
      total_24h = "24 h total"
    )
  ) +
  scale_fill_manual(
    values = c(
      "breeding" = "#F8766D",
      "winter" = "#619CFF"
    )
  )

plot_flight_distance_descriptive_daily

ggsave(
  file.path(results_dir, "plot_flight_distance_descriptive_daily.png"),
  plot_flight_distance_descriptive_daily,
  width = 15,
  height = 10,
  units = "cm",
  dpi = 600
)

```

# Flight height

### Data preparation and error calculation

Prepare the altitude dataset, identify phase-specific outliers, and estimate device- and acquisition-specific altitude error

```{r message=FALSE, warning=FALSE}

# List prepared per-individual datasets
files <- list.files(path = derived_data_dir, pattern = "Data_Prep\\.csv$")

# Store all valid altitude records across individuals
data_list <- list()

for (i in files) {
  
  x <- read.csv(file.path(derived_data_dir, i), sep = ",")
  
  x$UTC_datetime <- as.POSIXct(x$UTC_datetime, format = "%Y-%m-%d %H:%M:%OS", tz = "GMT")
  x <- x[order(x$UTC_datetime),]
  
  # Retain points previously classified as valid during the preparation workflow
  x <- x[is.na(x$filter) | x$filter == "OK", ]
  
  # Harmonise identifier types before binding all individuals
  x <- x %>%
    mutate(
      device_id = as.character(device_id),
      Individual_ID = as.character(Individual_ID)
    )
  
  # Remove records without behavioural classification
  x <- x[!is.na(x$behavior), ]
  
  data_list[[length(data_list) + 1]] <- x
}

# Merge all individuals and harmonise acquisition-mode labels
data_height_complete <- bind_rows(data_list) %>%
  mutate(
    acquisition_mode = dplyr::recode(
      acquisition_mode,
      "boost_start" = "boost",
      "boost_end"   = "boost",
      "burst_end"   = "burst"
    )
  ) %>%
  filter(!is.na(altitude_above_ground), !is.na(acquisition_mode))

# Inspect the stationary altitude distribution by device and acquisition mode
ggplot(
  data_height_complete[data_height_complete$behavior == "stationary" & !is.na(data_height_complete$altitude_above_ground), ],
  aes(x = altitude_above_ground, color = acquisition_mode)
) +
  geom_density(linewidth = 1) +
  scale_x_continuous(
    limits = c(-40, 40),
    breaks = seq(-40, 40, by = 5)
  ) +
  facet_wrap(~ gps_model) +
  labs(
    title = "Altitude Distribution by GPS Model (stationary)",
    x = "GPS Altitude (m)",
    y = "Density",
    color = "Acquisition mode"
  ) +
  theme_classic()

# Define empirical outlier limits separately for stationary and flight records
stationary_altitude_limits <- data_height_complete %>%
  filter(behavior == "stationary") %>%
  group_by(gps_brand, gps_model) %>%
  summarise(
    P2.5  = quantile(altitude_above_ground, 0.025, na.rm = TRUE),
    P97.5 = quantile(altitude_above_ground, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

flight_altitude_limits <- data_height_complete %>%
  filter(behavior == "flight") %>%
  group_by(gps_brand, gps_model, phen_phase) %>%
  summarise(
    P2.5  = quantile(altitude_above_ground, 0.025, na.rm = TRUE),
    P97.5 = quantile(altitude_above_ground, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# Flag stationary outliers using device-specific empirical limits
data_height <- data_height_complete %>%
  left_join(stationary_altitude_limits, by = c("gps_brand", "gps_model")) %>%
  mutate(
    outlier_height = case_when(
      behavior == "stationary" &
        (altitude_above_ground < P2.5 | altitude_above_ground > P97.5) ~ "yes",
      TRUE ~ "no"
    )
  ) %>%
  dplyr::select(-P2.5, -P97.5)

# Flag flight outliers using phase-specific empirical limits
data_height <- data_height %>%
  left_join(flight_altitude_limits, by = c("gps_brand", "gps_model", "phen_phase")) %>%
  mutate(
    outlier_height = case_when(
      behavior == "flight" &
        (altitude_above_ground < P2.5 | altitude_above_ground > P97.5) ~ "yes",
      TRUE ~ outlier_height
    )
  ) %>%
  dplyr::select(-P2.5, -P97.5)

# Prepare stationary data used to quantify GPS vertical error
stationary_height_data <- data_height %>%
  filter(behavior == "stationary", outlier_height == "no") %>%
  group_by(gps_brand, gps_model, acquisition_mode) %>%
  mutate(
    residual_to_median = altitude_above_ground - median(altitude_above_ground, na.rm = TRUE),
    abs_dev_from_median = abs(residual_to_median),
    abs_error = abs(altitude_above_ground)
  ) %>%
  ungroup()

# Summarise trueness, precision, and accuracy metrics for stationary fixes
device_error <- stationary_height_data %>%
  group_by(gps_brand, gps_model, acquisition_mode) %>%
  summarise(
    n_stationary_fixes = dplyr::n(),
    n_individuals = n_distinct(Individual_ID),
    trueness_mean_signed_error_m = mean(altitude_above_ground, na.rm = TRUE),
    trueness_median_signed_error_m = median(altitude_above_ground, na.rm = TRUE),
    precision_sd_m = sd(residual_to_median, na.rm = TRUE),
    precision_mean_abs_dev_m = mean(abs_dev_from_median, na.rm = TRUE),
    precision_median_abs_dev_m = median(abs_dev_from_median, na.rm = TRUE),
    precision_q95_abs_dev_m = quantile(abs_dev_from_median, 0.95, na.rm = TRUE),
    precision_rmse_m = sqrt(mean(residual_to_median^2, na.rm = TRUE)),
    accuracy_mean_abs_error_m = mean(abs_error, na.rm = TRUE),
    accuracy_median_abs_error_m = median(abs_error, na.rm = TRUE),
    accuracy_q95_abs_error_m = quantile(abs_error, 0.95, na.rm = TRUE),
    accuracy_rmse_m = sqrt(mean(altitude_above_ground^2, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(species = "Pterocles orientalis") %>%
  relocate(species, .before = 1)

tagList(
  h3("GPS vertical error metrics"),
  datatable(device_error, options = list(scrollX = TRUE, pageLength = 10))
)

write.csv(device_error, file.path(results_dir, "GPS_error_metrics.csv"), row.names = FALSE)

# Append device-specific error terms to each record
data_height <- data_height %>%
  left_join(device_error, by = c("gps_brand", "gps_model", "acquisition_mode"))

# Export the height dataset used in the altitude analyses
data_height <- data_height %>%
  dplyr::select(
    Individual_ID,
    UTC_datetime,
    phen_phase,
    tod,
    acquisition_mode,

    behavior,
    ground_elevation,
    altitude_above_ground,
    gps_model,
    gps_model_full,
    outlier_height,
    n_stationary_fixes,
    n_individuals,
    trueness_median_signed_error_m,
    precision_sd_m,
    
    satcount,
    Altitude_m,
    Altitude_ellipsoid_m,
    speed_km_h,
    Course,
    Positioning.mode,
    HDOP,
    VDOP
  )

write.csv(data_height, file.path(results_dir, "Flight_height_data.csv"), row.names = FALSE)

```

### Bootstrap

Propagate vertical GPS error by repeatedly correcting altitude above ground using device-specific bias and residual error estimates

```{r message=FALSE, warning=FALSE}

plan(multisession)

# Remove systematic bias and add one random draw from the residual error distribution
correct_altitude <- function(altitude, bias_m, precision_sd_m) {
  mapply(function(alt, bias, sd_val) {
    alt_corr <- alt - bias
    alt_corr + rnorm(1, mean = 0, sd = sd_val)
  }, altitude, bias_m, precision_sd_m)
}

# Generate bootstrap-corrected altitudes within each individual × phase × tod group
bootstrap_corrected_altitudes <- function(df, n_boot = 100) {
  purrr::map(1:n_boot, function(i) {
    correct_altitude(
      altitude = df$altitude_above_ground,
      bias_m = df$trueness_median_signed_error_m,
      precision_sd_m = df$precision_sd_m
    )
  }) %>%
    dplyr::bind_cols()
}

# Retain only flight records used in the height analysis
flight_data <- data_height %>%
  filter(behavior == "flight", outlier_height == "no") %>%
  filter(!is.na(Individual_ID))

# Apply bootstrap correction separately within each phase × tod × individual group
flight_groups_by_phase_tod_id <- flight_data %>%
  group_by(phen_phase, tod, Individual_ID) %>%
  group_split()

bootstrap_altitudes_list <- future_map(
  flight_groups_by_phase_tod_id,
  bootstrap_corrected_altitudes,
  n_boot = 100,
  .options = furrr_options(seed = 1234)
)

# Reshape the bootstrap output for downstream summaries
bootstrap_altitudes_long <- map2_dfr(
  bootstrap_altitudes_list,
  flight_groups_by_phase_tod_id,
  function(altitudes_df, original_df) {
    altitudes_df %>%
      mutate(row_id = 1:n()) %>%
      pivot_longer(
        cols = -row_id,
        names_to = "bootstrap",
        values_to = "corrected_altitude"
      ) %>%
      mutate(
        phen_phase = original_df$phen_phase[1],
        tod = original_df$tod[1],
        Individual_ID = original_df$Individual_ID[1]
      )
  }
)

plan(sequential)

```

### Descriptive summaries of corrected flight height by phenological phase and daytime period

```{r}

# Count data support directly from the non-bootstrapped flight dataset
phase_levels <- c("breeding", "winter")
phase_levels <- phase_levels[phase_levels %in% unique(flight_data$phen_phase)]

flight_height_support_tod <- flight_data %>%
  filter(phen_phase != "other") %>%
  group_by(phen_phase, tod) %>%
  summarise(
    n_individuals = n_distinct(Individual_ID),
    n_fixes = dplyr::n(),
    .groups = "drop"
  )

flight_height_support_phase <- flight_data %>%
  filter(phen_phase != "other") %>%
  group_by(phen_phase) %>%
  summarise(
    n_individuals = n_distinct(Individual_ID),
    n_fixes = dplyr::n(),
    .groups = "drop"
  ) %>%
  mutate(tod = "all") %>%
  relocate(tod, .after = phen_phase)

# Summarise the bootstrap distribution of corrected flight height by phenological phase and daytime period
flight_height_summary_tod <- bootstrap_altitudes_long %>%
  filter(phen_phase != "other") %>%
  group_by(bootstrap, phen_phase, tod) %>%
  summarise(
    mean_height_boot_m = mean(corrected_altitude, na.rm = TRUE),
    median_height_boot_m = median(corrected_altitude, na.rm = TRUE),
    q25_height_boot_m = quantile(corrected_altitude, 0.25, na.rm = TRUE),
    q75_height_boot_m = quantile(corrected_altitude, 0.75, na.rm = TRUE),
    sd_height_boot_m = sd(corrected_altitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(phen_phase, tod) %>%
  summarise(
    mean_height_m = median(mean_height_boot_m, na.rm = TRUE),
    mean_height_lcl_m = quantile(mean_height_boot_m, 0.025, na.rm = TRUE),
    mean_height_ucl_m = quantile(mean_height_boot_m, 0.975, na.rm = TRUE),
    median_height_m = median(median_height_boot_m, na.rm = TRUE),
    median_height_lcl_m = quantile(median_height_boot_m, 0.025, na.rm = TRUE),
    median_height_ucl_m = quantile(median_height_boot_m, 0.975, na.rm = TRUE),
    q25_height_m = median(q25_height_boot_m, na.rm = TRUE),
    q75_height_m = median(q75_height_boot_m, na.rm = TRUE),
    sd_height_m = median(sd_height_boot_m, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(flight_height_support_tod, by = c("phen_phase", "tod"))

# Summarise the bootstrap distribution of corrected flight height by phenological phase only
flight_height_summary_phase <- bootstrap_altitudes_long %>%
  filter(phen_phase != "other") %>%
  group_by(bootstrap, phen_phase) %>%
  summarise(
    mean_height_boot_m = mean(corrected_altitude, na.rm = TRUE),
    median_height_boot_m = median(corrected_altitude, na.rm = TRUE),
    q25_height_boot_m = quantile(corrected_altitude, 0.25, na.rm = TRUE),
    q75_height_boot_m = quantile(corrected_altitude, 0.75, na.rm = TRUE),
    sd_height_boot_m = sd(corrected_altitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(phen_phase) %>%
  summarise(
    mean_height_m = median(mean_height_boot_m, na.rm = TRUE),
    mean_height_lcl_m = quantile(mean_height_boot_m, 0.025, na.rm = TRUE),
    mean_height_ucl_m = quantile(mean_height_boot_m, 0.975, na.rm = TRUE),
    median_height_m = median(median_height_boot_m, na.rm = TRUE),
    median_height_lcl_m = quantile(median_height_boot_m, 0.025, na.rm = TRUE),
    median_height_ucl_m = quantile(median_height_boot_m, 0.975, na.rm = TRUE),
    q25_height_m = median(q25_height_boot_m, na.rm = TRUE),
    q75_height_m = median(q75_height_boot_m, na.rm = TRUE),
    sd_height_m = median(sd_height_boot_m, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(tod = "all") %>%
  relocate(tod, .after = phen_phase) %>%
  left_join(flight_height_support_phase, by = c("phen_phase", "tod"))

# Combine TOD-specific and phase-level summaries
flight_height_summary <- bind_rows(
  flight_height_summary_tod,
  flight_height_summary_phase
) %>%
  mutate(
    species = "Pterocles orientalis",
    phen_phase = factor(phen_phase, levels = phase_levels),
    tod = factor(tod, levels = c("dawn", "day", "dusk", "night", "all"))
  ) %>%
  arrange(phen_phase, tod) %>%
  relocate(species, .before = 1)

tagList(
  h3("Descriptive corrected flight height by phenological phase and daytime period"),
  datatable(
    flight_height_summary,
    options = list(scrollX = TRUE, scrollY = TRUE, pageLength = 12)
  )
)

write.csv(
  flight_height_summary,
  file.path(results_dir, "Flight_height_summary.csv"),
  row.names = FALSE
)


# Plot descriptive median flight height with bootstrap percentile intervals
plot_flight_height_descriptive <- flight_height_summary %>%
  ggplot(aes(x = tod, y = median_height_m, color = phen_phase)) +
  geom_errorbar(
    aes(ymin = median_height_lcl_m, ymax = median_height_ucl_m),
    position = position_dodge(width = 0.8),
    width = 0.2,
    colour = "black"
  ) +
  geom_point(
    position = position_dodge(width = 0.8),
    size = 2.5
  ) +
  facet_wrap(~ phen_phase) +
  labs(
    x = "Daytime period",
    y = "Corrected median flight height (m)"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(
    values = c(
      "breeding" = "#F8766D",
      "winter" = "#619CFF"
    )
  )

plot_flight_height_descriptive

ggsave(
  file.path(results_dir, "plot_flight_height_descriptive.png"),
  plot_flight_height_descriptive,
  width = 15,
  height = 10,
  units = "cm",
  dpi = 600
)

```

# Session information

R and package versions used to render this analytical workflow.

```{r session-information}
sessionInfo()
```
